---
title: "EDDA - Assignment 2 - Group 77"
subtitle: "Dante de Lang, Ignas Krikštaponis and Kamiel Gülpen"
output:
  pdf_document: default
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
library(tidyverse)
library(rstudioapi)
library(lme4)

knitr::opts_chunk$set(echo = TRUE)
setwd(dirname(getActiveDocumentContext()$path))
# round numbers to 3 digits
options(digits = 3)
```

# Exercise 1
If left alone bread will become moldy, rot or decay otherwise. To investigate the influence of temperature and humidity on this process, the time to decay was measured for 18 slices of white bread, which were placed in 3 different environments and humidified or not. The data are given in the file bread.txt, with the first column time to decay in hours, the second column the environment (cold, warm or intermediate temperature) and the third column the humidity.

**a)** The 18 slices came from a single loaf, but were randomized to the 6 combinations of conditions. Present an R-code for this randomization process.

```{r}
# randomization code
breads <- 18; environments <- 3; humidities <- 2;
as_tibble(
      cbind(bread = 1:breads,
      environment = sample(rep(c("cold", "warm", "intermediate"),breads/environments)),
      humidity = rep(c("dry", "wet"), breads/humidities))
)
```

**b)** Make two boxplots of hours versus the two factors and two interaction plots (keeping the two factors fixed in turn).

```{r}
# read data
data <- read.table(file="data/bread.txt",header=TRUE)
data$environment <- as.factor(data$environment); data$humidity <- as.factor(data$humidity)
attach(data)
# boxplots
par(mfrow=c(1,2))
boxplot(hours~environment)
boxplot(hours~humidity)
# interaction plots
interaction.plot(environment,humidity,hours)
interaction.plot(humidity,environment,hours) 
```

**c)** Perform an analysis of variance to test for effect of the factors temperature, humidity, and their interaction. Describe the interaction effect in words.

```{r}
aov <- lm(hours ~ environment * humidity, data = data)
anova(aov)
summary(aov)
```

From the above p-values we can conclude that:
* that the levels of environment are associated with significant different decay hours.
* that the levels of humidity are associated with significant different decay hours.
* the relationships between decay hours and environment depends on the humidity level.

**d)** Which of the two factors has the greatest (numerical) influence on the decay? Is this a good question?

**e)** Check the model assumptions by using relevant diagnostic tools. Are there any outliers?
```{r}
par(mfrow=c(1,2))
plot(aov, 2);plot(aov, 1)
```

The above normality and residuals diagnostics signal that there are outliers. By removing them we would be able to satisfy the normality assumptions required for the two-way ANOVA test.

# Exercise 2

A researcher is interested in the time it takes a student to find a certain product on the internet using a search engine. There are three different types of interfaces with the search engine and especially the effect of these interfaces is of importance. There are five different types of students, indicating their level of computer skill (the lower the value of this indicator, the better the computer skill of the corresponding student). Fifteen students are selected; three from each group with a certain level of computer skill. The data is given in the file search.txt. Assume that the experiment was run according to a randomized block design which you make in a). (Beware that the levels of the factors are coded by numbers.)

**a** Number the selected students 1 to 15 and show how (by using R) the students could be randomized to the interfaces in a randomized block design.
```{r}

```

**b**
```{r}
# read data
data <- read.table(file="data/search.txt",header=TRUE)
data$skill <- as.factor(data$skill); data$interface <- as.factor(data$interface)
# perform ANOVA
aov <- lm(time ~ interface + skill, data = data)
anova(aov)
# summary table
summary(aov)
```

**c)** Check the model assumptions by using relevant diagnostic tools.
```{r}
par(mfrow=c(1,2))
plot(aov, 2);plot(aov, 1)
```
There are outliers therefore it would be best to use a different test or remove them.

**d)** Perform the Friedman test to test whether there is an effect of interface.
```{r}
attach(data)
friedman.test(time, interface, skill)
```
According to the Friedman test there is an effect of the interface.

**e)** Test the null hypothesis that the search time is the same for all interfaces by a one-way ANOVA test, ignoring the variable skill. Is it right/wrong or useful/not useful to perform this test on this dataset?

```{r}
one_aov <- lm(time ~ interface, data = data)
anova(one_aov)
```

From the test above, we could conclude that the interface has no siginificant affect on the time taken to find a certain product. However as we have multiple factors here, one-way ANOVA is not appropriate here as it will ignore the effects the ignored factor could have.

# Excercise 3

In a study on the effect of feedingstuffs on lactation a sample of nine cows were fed with two types of food, and their milk production was measured. All cows were fed both types of food, during two periods, with a neutral period in-between to try and wash out carry-over effects. The order of the types of food was randomized over the cows. The observed data can be found in the file cow.txt, where A and B refer to the types of feedingstuffs.

***a)*** Test whether the type of feedingstuffs influences milk production using an ordinary "fixed effects" model, fitted with lm. Estimate the difference in milk production.

```{r}
# read data
data <- read.table(file="data/cow.txt",header=TRUE)
data$treatment <- as.factor(data$treatment); data$order <- as.factor(data$order)
data$id <- as.factor(data$id); data$per <- as.factor(data$per)

# perform fixed effects model analysis
fixed_aov <- lm(milk ~ order + id + per + treatment, data = data)
summary(fixed_aov)
```

***b)***
```{r warning=FALSE}
attach(data)
mixed_avo <- lmer(milk ~ treatment + order + per + (1|id),REML=FALSE)
mixed_avo_1 <- lmer(milk ~ order + per + (1|id),REML=FALSE)
anova(mixed_avo_1, mixed_avo)
```

***c)***
```{r, warning=FALSE}
attach(data)
t.test(milk[treatment=="A"],milk[treatment=="B"],paired=TRUE)
```

# Excercise 4

Stochastic models for word counts are used in quantitative studies on literary styles. Statistical analysis of the counts can, for example, be used to solve controversies about true authorships. Another example is the analysis of word frequencies in relation to Jane Austen's novel Sanditon. At the time Austen died, this novel was only partly completed. Austen, however, had made a summary for the remaining part. An admirer of Austen's work finished the novel, imitating Austen's style as much as possible. The file austen.txt contains counts of different words in some of Austen's novels: chapters 1 and 3 of Sense and Sensibility (stored in the Sense column), chapters 1, 2 and 3 of Emma (column Emma), chapters 1 and 6 of Sanditon (both written by Austen herself, column Sand1) and chapters 12 and 24 of Sanditon (both written by the admirer, Sand2).

***a)*** Discuss whether a contingency table test for independence or for homogeneity is most appropriate here.
```{r}
data <- read.table(file="data/austen.txt",header=TRUE)
```
Homogeneity is the most appropretiate test here since we want to see if all chapters are homogeneous and if they are not, then the admirer had a different writting stile.

***b)*** Using the given data set, investigate whether Austen herself was consistent in her different novels. Where are the main inconsistencies?
```{r}
only_austen <- data %>% 
  select(Sense, Emma, Sand1)
test <- chisq.test(only_austen); test
test$residuals
```
From the p-value it seems that there are no significant difference between word counts in Austen novels' chapters.

***c)*** Was the admirer successful in imitating Austen's style? Perform a test including all data. If he was not successful, where are the differences?
```{r}
test <- chisq.test(data); test
test$residuals
```
From the p-value above we can conclude that there are significant word count differences between chapters, therefore the admirer fail to mimic Austen's style. The main differences lay in word an and that.

# Excercise 5

***a)*** The data in expensescrime.txt were obtained to determine factors related to state expenditures on criminal activities (courts, police, etc.) The variables are: state (indicating the state in the USA), expend (state expenditures on criminal activities in $1000), bad (crime rate per 100000), crime (number of persons under criminal supervision), lawyers (number of lawyers in the state), employ (number of persons employed in the state) and pop (population of the state in 1000). In the regression analysis, take expend as response variable and bad, crime, lawyers, employ and pop as explanatory variables.

***a)*** Make some graphical summaries of the data. Investigate the problem of potential and in
uence points, and the problem of collinearity.

```{r}
data <- read.table(file="data/expensescrime.txt",header=TRUE)
plot(data[,c(3, 4, 5, 6, 7, 2)])
par(mfrow=c(2,3))
for (i in c(3, 4, 5, 6, 7, 2)) hist(data[,i],main=names(data)[i])
```

***b)*** Fit a linear regression model to the data. Use both the step-up and the step-down method to find the best model. If step-up and step-down yield two different models, choose one and motivate your choice.

```{r}
# step down model
down_lm_1  <- lm(expend ~ bad + crime + lawyers + employ + pop ,data = data); summary(down_lm_1)
down_lm_2  <- lm(expend ~ bad + lawyers + employ + pop ,data = data); summary(down_lm_2)
# step up model
```

***c)*** Check the model assumptions by using relevant diagnostic tools.

```{r}
par(mfrow=c(1,2))
plot(down_lm_2, 2);plot(down_lm_2, 1)
```