---
title: "EDDA - Assignment 2 - Group 77"
subtitle: "Dante de Lang, Ignas Krikštaponis and Kamiel Gülpen"
output:
  pdf_document: default
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
library(tidyverse)
library(rstudioapi)

knitr::opts_chunk$set(echo = TRUE)
setwd(dirname(getActiveDocumentContext()$path))
# round numbers to 3 digits
options(digits = 3)
```


# Exercise 1
Moldy bread If left alone bread will become moldy, rot or decay otherwise.  To investigate the influence of temperature and  humidity  on  this  process,  the  time  to  decay  was  measured  for  18  slices  of  white  bread,  which  were placed in 3 different environments and humidified or not.  The data are given in the file bread.txt, with the first column time to decay in hours, the second column the environment (cold, warm or intermediate temperature) and the third column the humidity.


**a)**The 18 slices came from a single loaf, but were randomized to the 6 combinations of conditions.  Present an R-code for this randomization process.
```{r}
data <- read.table(file="data/bread.txt",header=TRUE)
I=3; J=2; N=3
env = rep(c("cold","intermediate","warm"), each=N*J)
hum = rep(c("dry", "wet"),each = N*I)
cbind(env, hum, sample(1:(N*I*J)))
```

**b)**Make two boxplots of hours versus the two factors and two interaction plots (keeping the two factors fixed in turn).
```{r}
par(mfrow=c(1,2))
attach(data)
boxplot(hours~environment)
boxplot(hours~humidity)
interaction.plot(humidity,environment,hours)
interaction.plot(environment,humidity,hours)

```


**c)**Perform  an  analysis  of  variance  to  test  for  effect  of  the  factors temperature,humidity,  and  the interaction.  Describe the interaction effect in words.
```{r}

data$environment=as.factor(data$environment)
data$humidity=as.factor(data$humidity)
dataaov=lm(hours~humidity*environment)
anova(dataaov)

summary(dataaov)

# Without interaction
#data$humidity=as.factor(data$humidity)
#data$environment=as.factor(data$environment)
#dataaov=lm(hours~humidity+environment,data=data)
#anova(dataaov)

```
When looking at the two-way anova model we see that it consists of the following terms: Y_{ijk} = $\mu_{ij}$ + e_{ijk} = $\mu + alpha_{i}$ + $\beta_{j}$ + $\gamma_{ij}$ + $e_{eijk}$ We decompose the formula it this way such that $\mu$ is the overall mean, $\alpha_{i}$ and $\beta_{j}$ are the main effect of level i and j of the first factor and second factor respectively and $\gamma_{ij}$ the interaction effect. 

In order to test the effect of the temperature,humidity,  and  the interaction we set up 3 hypotheses which are:
$H_{AB}$: $\gamma_{ij}$ = 0 for every (i, j) (no interactions between factor A and B)

$H_{A}$: $\alpha_{i}$ = 0 for every i (no main effect of factor A)

$H_{B}$:$\beta_{j}$ = 0 for every j (no main effect of factor B)

We use the test statistics $F_{AB}$ for $H_{AB}$, $F_{A}$ for $H_{A}$ and $F_{B}$ for $H_{B}$ where F is the F-distribution. 

To see if the Hypotheses can be rejected we want to look at the probability that P(F>$f_{AB}$), P(F>$f_{A}$) and P(F>$f_{B}$), the bigger the F value the lower the probability that the Hypothesis lays under a F-distribution and therefore the Hypothesis can be rejected.

We see that the humidity has a p-value of 4.3e-06, environment a p-value of 2.5e-10 and the interaction between the two (humidity:environment) shows a p-value of 3.7e-07. This means that humidity, environment and the interaction effect between humidity and environment have a significant influence on the hours, which means we can reject $H_{A}$, $H_{B}$ and $H_{AB}$.

The interaction effect looks at the difference of differences, for example: it looks at the difference in hours for environment = cold and environment = warm for humidity = wet. Then it looks the difference between environment = cold and environment = warm for humidity = dry. It then looks at the difference betwen those differences and when this difference is high it shows that there is indeed interaction.

**d)** Which of the two factors has the greatest (numerical) influence on the decay?  Is this a good question?

When we want to know which factor has the greatest influence we want to use the additive model as used above. This shows a p-value of 0.026 for humidity and a p-value of 3.7e-05 for environemnt. This means that the environment has the greatest influence. 


**e)**  Check the model assumptions by using relevant diagnostic tools.  Are there any outliers?

```{r}
par(mfrow=c(1,2))
dataaov2=lm(hours~humidity*environment,data=data); 
shapiro.test(residuals(dataaov2))
plot(dataaov2, 1)
plot(dataaov2, 2)
```
The qqplot shows a somewhat linear line which means that based on the qqplot we can state that the data is normally distributed. Furthermore we used a Shapiro-Wilks test to see if the test can back this assumption. The Shapiro-Wils test showed a p-value of 0.2 which means that the data is normally distributed. There is also looked at the spread of the residuals, which showed that there are three outliers which are number 5, 7 and 8 which can be observed in both plot.


# Exercise 2

A researcher is interested in the time it takes a student to find a certain product on the internet using a search engine. There are three different types of interfaces with the search engine and especially the effect of these interfaces is of importance. There are five different types of students, indicating their level of computer skill (the lower the value of this indicator, the better the computer skill of the corresponding student). Fifteen students are selected; three from each group with a certain level of computer skill. The data is given in the file search.txt. Assume that the experiment was run according to a randomized block design which you make in a). (Beware that the levels of the factors are coded by numbers.)

**a)** Number the selected students 1 to 15 and show how (by using R) the students could be randomized to the interfaces in a randomized block design.

```{r}
data_search <- read.table(file="data/search.txt",header=TRUE)

interface <- factor(rep(c("1","2","3"),each = 5))
skill <- factor(rep(c("1","2","3","4","5"),times = 3))
students <- c(1:15)
block <- data.frame(students,skill,interface)

block

```

**b)** Test the null hypothesis that the search time is the same for all interfaces. What type of interface does require the longest search time? For which combination of skill level and type of interface is the search time the shortest? Estimate the time it takes a typical user of skill level 3 to find the product on the website if the website uses interface 3.

```{r}
attach(data_search)
aovsearch = lm(time~interface+skill)
anova(aovsearch)

boxplot(time~interface) # Interface 3 has the longest search time

interaction.plot(skill,interface,time) # Skill 2 and interface 1 is the fastest

summary(aovsearch) # Estimate interface 3 = 4.46, skill 3 = 3.03, so 3-3 gives:

(4.46+3.03)/2 # 3.75 seconds ????

```

**c)** Check the model assumptions by using relevant diagnostic tools.

```{r}
par(mfrow=c(1,2))
qqnorm(residuals(aovsearch))
plot(fitted(aovsearch),residuals(aovsearch))
```

**d)** Perform the Friedman test tot test whether there is an effect of interface.

```{r}
friedman.test(time,interface,skill)
```

**e)**Test the null hypothesis that the search time is the same for all interfaces by a one-way ANOVA test, ignoring the variable skill. Is it right/wrong or useful/not useful to perform this test on this dataset?

```{r}
attach(data_search)

aovsearch = lm(time~interface)
anova(aovsearch)
```


# Exercise 4

Stochastic models for word counts are used in quantitative studies on literary styles.  Statistical analysis of the counts can, for example, be used to solve controversies about true author ships.  Another example is the analysis of word frequencies in relation to Jane Austen’s novel Sanditon.  At the time Austen died, this novel was only partly completed.  Austen, however, had made a summary for the remaining part.  An admirer of Austen’s work finished the novel, imitating Austen’s style as much as possible.  The file austen.txt contains counts of different words in some of Austen’s novels:  chapters 1 and 3 of Sense and Sensibility(stored in the Sense column), chapters 1, 2 and 3 of Emma(column Emma), chapters 1 and 6 of Sanditon(both written by Austen herself, column Sand1) and chapters 12 and 24 of Sanditon(both written by the admirer,Sand2)

**a)**
Discuss whether a contingency table test for independence or for homogeneity is most appropriate here.

The contingency table test for homogeneity is appropriate because we want to know if the fan writer imitates Austen in a good way. This means that we want to test whether or not the different columns of data in the table come from the same population (writer) or not, which would be the case it the fan imitated Austen correctly. The H0 of the contingency table test for homogeneity states that the distribution of the words is the same for the stories.

**b)** Using  the  given  data  set,  investigate  whether  Austen  herself  was  consistent  in  her  different  novels. Where are the main inconsistencies?

```{r}
data=read.table(file="data/austen.txt",header=TRUE)
austen = data[,1:3]
z = chisq.test(austen)
z
residuals(z)
```

She is not inconsistent as the p-value is above 0.05. This means that we cannot reject the H0. She does however have some main inconsistency, which where the words "a", "that" and "without". As can be seen in the residual tanle above.

```{r}
z = chisq.test(data)
z
residuals(z)

```
The fan is inconsistent as the p-value of the test is below 0.05. Therefore we have to reject the H0 and accept that the distribution of the words in the stories are not the same. Because Austen herself did not have this inconsistency we can say that the inconsistency is caused by the fan writer. The main inconsistencies were for the words  "that" and "an". As can be seen in the residual tanle above.

