---
title: "EDDA - Assignment 1 - Group 77"
subtitle: "Dante de Lang, Ignas Krikštaponis and Kamiel Gülpen"
output:
  pdf_document: default
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
library(tidyverse)
library(rstudioapi)

knitr::opts_chunk$set(echo = TRUE)
setwd(dirname(getActiveDocumentContext()$path))
# round numbers to 3 digits
options(digits = 3)
```

# Exercise 1
The data set birthweight.txt contains the birthweights of 188 newborn babies.  We are interested in finding the underlying (population) mean $\mu$ of birthweights.

**a)** Check normality of the data.  Compute a point estimate for mu.  Derive, assuming normality (irrespective of your conclusion about normality of the data), a bounded 90% confidence interval for $\mu$.


To check normality for the data we use a qqplot, historgram, box plot and Shapiro-Wilks test. 
```{r}
par(mfrow=c(1,3))
data <- read.table(file="data/birthweight.txt",header=TRUE)
data
hist(data$birthweight)
shapiro.test(data$birthweight)
qqnorm(data$birthweight)



error = qnorm(0.95)*(sd(data$birthweight)/sqrt(length((data$birthweight))))

ci = c(mean(data$birthweight) - error, mean(data$birthweight) + error)
ci
```
```{r}
t.test(data$birthweight, alternative = "greater", mu=2800)
```
```{r}
data <- read.table(file="data/run.txt",header=TRUE)
data
bef_limo = data[ which(data$drink=='lemo'),]$before 
bef_energy =data[ which(data$drink=='energy'),]$before 
aft_limo = data[ which(data$drink=='lemo'),]$after
aft_energy =  data[ which(data$drink=='energy'),]$after

aft_limo
lemo = c()
energy = c()
for (i in 1:length(aft_limo))
  {y = abs(bef_energy[i] - aft_energy[i]); x = abs(bef_limo[i] - aft_limo[i]); lemo = c(lemo,x );energy = c(energy, y); y}

lemo
energy

hist(lemo)
hist(energy)

wilcox.test(lemo, energy)
ks.test(lemo,energy)
```
